CUDA-accelerated Genetic Feedforward-ANN Training for Data Mining

### 2010-04-22 doc/report.tex
Small tweaks in wording
git-svn-id: http://svn.scesoc.ca/evolution/trunk@115 fee98b4c-d1c3-4d34-9525-a468159e4a48

% ----------------------------------------------------------------------------
\subsection{Preliminary Data Analysis} \label{prelim}
% ----------------------------------------------------------------------------
To get a first idea of the separability of the data, we took an approach similar to Fisher's linear discriminant. This discriminant finds the 1D projection of the data which maximizes class separation, defined as the ratio of between-class variance and within-class variance. The problem of finding Fisher's linear discriminant can be seen as a generalized eigenvalue problem where the eigenvector with the large eigenvalue is the desired projection \cite[\S 3.8.2]{Duda}.
To gain more insight into the higher-dimensional properties of the data, and because 2D visualization is practical, we instead used the eigenvectors with the \emph{two} highest eigenvalues and formed a 2D projection of the data (Figure~\ref{fig:fishers-2d}). The resulting four approximately-Gaussian clusters guided our decision to use a radial basis function network with four Gaussian hidden nodes.

@@

% ----------------------------------------------------------------------------
\subsection{Data Preprocessing Steps} \label{preprocessing}
% ----------------------------------------------------------------------------
In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the data set into numeric values.  The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative classifications observed given the nominal value.  As a result, meaningful numeric values can be extracted from nominal values without domain knowledge relating to the values. This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.  The logarithm of the ratio of all positive classifications to all negative classifications is attributed to nominal values which do not appear sufficiently in the data set and for values which appear only in blind test data sets; this prevents the classifier from attributing significant values to nominal values for which little or no inference can be done.

@@

To avoid this issue, we standardize the data before training and before classification. For each feature individually, we estimate the 10\textsuperscript{th} and 90\textsuperscript{th} percentiles of the feature values. Then we shift and scale the values such that 80\% of the resulting values are found between -1 and 1. This standardization method was chosen for its robustness against outliers. After standardization, the range of inputs to \texttt{expf} is narrow enough to always produce a valid FP value at the output. The standardization does not affect the classification itself because there are scaling and shifting parameters within the ANN (input to hidden edge weights and hidden center parameters, respectively) which are already part of the search space.

@@

Our GA begins by generating an initial population of random candidate ANNs, each of which is a set of parameters describing one ANN. Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations. Then, the following steps are executed in sequence for each generation of the GA:

@@

% ----------------------------------------------------------------------------
\subsection{Classification Performance} \label{results}
% ----------------------------------------------------------------------------
The classifier was trained over 12000 generations; the fitness metric during this training was lift at 20\% calculated on a training set which consisted of 70\% of the data set, chosen at random during initialization.  After each generation, the candidate ANN with the highest fitness value was used to classify 100 random subsets of the hold-out data set (each consisting of exactly one half of the hold-out data set), and the average of the lift at 20\% of these classifications was recorded.  Figure~\ref{fig:evolution-lift} demonstrates experimental results across 12,000 generations of population 50. The final estimated lift is 4.51 out of a maximum lift of 5.

@@

When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35th in the world for classification of this data set. Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking. The discrepency between the estimated performance and actual performance on the blind test data set can be attributed to the following:
\begin{itemize}
	\item Discrepency in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
	\item Discrepency in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets. The blind test data set was known to have the same class balance as the given data set. Therefore, it is possible that the hold-out data set used to collect experimental results had a class balance which did not reflect the blind test data set.
	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification, however, the pre-processing step which converts nominal values into numeric values accounts for values which have not been seen previously.
\end{itemize}

>>>>

% ----------------------------------------------------------------------------
\subsection{Preliminary Data Analysis} \label{prelim}
% ----------------------------------------------------------------------------
To get a first idea of the separability of the data, we took an approach similar to Fisher's linear discriminant. This discriminant finds the 1D projection of the data which maximizes class separation, defined as the ratio of between-class variance and within-class variance. The problem of finding Fisher's linear discriminant can be seen as a generalized eigenvalue problem where the eigenvector with the largest corresponding eigenvalue is the desired projection \cite[\S 3.8.2]{Duda}.
To gain more insight into the higher-dimensional properties of the data, and because 2D visualization is practical, we instead used the eigenvectors with the \emph{two} highest eigenvalues and formed a 2D projection of the data (Figure~\ref{fig:fishers-2d}). The resulting four approximately-Gaussian clusters guided our decision to use a radial basis function network with four Gaussian hidden nodes.

@@

% ----------------------------------------------------------------------------
\subsection{Data Preprocessing Steps} \label{preprocessing}
% ----------------------------------------------------------------------------
In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab separated values (TSV) file, converts nominal feature values in the data set into numeric values.  The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative classifications observed given the nominal value.  This is inspired from the discriminant function of a Bayesian classifier trained only on that particular feature. As a result, meaningful numeric values can be extracted from nominal values without domain knowledge relating to the values. This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.  The logarithm of the ratio of all positive classifications to all negative classifications is assigned to nominal values which do not appear sufficiently in the data set and for values which appear only in blind test data sets; this prevents the classifier from assigning significant values to nominal values for which little or no inference can be done.

@@

To avoid this issue, we standardize the data before training and before classification. For each feature individually, we estimate the 10\textsuperscript{th} and 90\textsuperscript{th} percentiles of the feature values. We then shift and scale the values such that 80\% of the resulting values are in the range $(-1, 1)$. This standardization method was chosen for its robustness against outliers. After standardization, the range of inputs to \texttt{expf} is narrow enough to always produce a valid FP value at the output. The standardization does not affect the classification itself because there are scaling and shifting parameters within the ANN (input to hidden edge weights and hidden center parameters, respectively) which are already part of the search space.

@@

Our GA begins by generating an initial population of random candidate ANNs, each of which is a set of parameters describing one ANN. Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations. The following steps are then executed in sequence for each generation of the GA:

@@

% ----------------------------------------------------------------------------
\subsection{Classification Performance} \label{results}
% ----------------------------------------------------------------------------
The classifier was trained over 12000 generations; the fitness metric during this training was lift at 20\% calculated on a training set which consisted of 70\% of the data set, chosen at random during initialization.  After each generation, the candidate ANN with the highest fitness value was used to classify 20 random subsets of the hold-out data set (each consisting of exactly one half of the hold-out data set), and the average of the lift at 20\% of these classifications was recorded.  Figure~\ref{fig:evolution-lift} demonstrates experimental results across 12,000 generations of population 50. The final estimated lift is 4.51 out of a maximum lift of 5.

@@

When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35\textsuperscript{th} in the world for classification of this data set. Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking. The discrepancy between the estimated performance and actual performance on the blind test data set can be attributed to the following:
\begin{itemize}
	\item Discrepancy in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
	\item Discrepancy in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets. The blind test data set was known to have the same class balance as the given data set. Therefore, it is possible that the hold-out data set used to collect experimental results had a class balance which did not reflect the blind test data set.
	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification. However, the pre-processing step which converts nominal values into numeric values accounts for values which are not previously seen.
\end{itemize}
